Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-23 trunk/paper/drinking.tex

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. The second common type of abnormality are
abnormalities based on process participants in relation to the
duration of the process. These include all kinds of {\em rates} such
as {\em mass flow rate}, {\em energy flow rate} and {\em velocity}
(the rate of change of position). In each of these cases, an object
participates in a process and a quality (or change of quality) of that
object throughout the duration of the process is considered to form a
new quality. If the process has participants that are distinguished
into {\em inputs} and {\em outputs}, then a recurring pattern is that
the amount of inputs or outputs that participate in the process can be
{\em increased} or {\em decreased}.  Finally, some objects may be
divided into stages during which particular states of affairs obtain,
and a process may be abnormal in that these states of affairs do not
obtain at a particular stage. Notably, at the beginning and the end of
a process, pre- and post-conditions may obtain that are abnormally
changed in a process. For example, {\em Aneuploidy} -- an abnormality
during cell division at which the chromosomes do not separate properly
between the two cells -- may be considered the result of such an
abnormality.

@@

There are further types of deviation from a distribution pattern. For
example, a kind of process that is normally {\em rhythmic} can be
abnormal in that it is {\em arrhythmic}. A typical example of this
kind of process is {\em Heart beating} ({\tt GO:0060047}), in which
{\em Cardiac muscle contraction} ({\tt GO:0060048}) processes occur in
a rhythmic pattern. In {\em Cardiac dysrhythmia}, however, {\em
  Cardiac muscle contraction} processes occur arrhythmically, and we
consider this to be an abnormality of the regulation of {\em Cardiac
  muscle contraction}.

A second kind of regulatory abnormalities are related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of {\em Regulation of apoptosis}
in which apoptosis is induced later than normal.

>>>>

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. For example, a part of an organism may
participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
lasts longer than normal, i.e., the organism has an {\em Increased
  duration of inflammatory response} phenotype. We define such a
phenotype as a phenotype of an organism which has a part that
participates in {\em Inflammatory response}, and this {\em
  Inflammatory response} process has an {\em Increased duration} ({\tt
  PATO:0000498}).

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em mass flow rate}, {\em
  energy flow rate} and {\em velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} can be defined as an increased amount of inputs or an
increased amount of outputs of a {\em cytoplasmic streaming} process.

Finally, some objects may be divided into stages during which
particular states of affairs obtain, and a process may be abnormal in
that these states of affairs do not obtain at a particular
stage. Notably, at the beginning and the end of a process, pre- and
post-conditions may obtain that are abnormally changed in a
process. For example, {\em Aneuploidy} -- an abnormality during cell
division at which the chromosomes do not separate properly between the
two cells -- may be considered the result of such an abnormality.

@@

There are further types of deviation from a distribution pattern. For
example, a kind of process that is normally {\em rhythmic} can be
abnormal in that it is {\em arrhythmic}. A typical example of this
kind of process is {\em Heart beating} ({\tt GO:0060047}), in which
{\em Cardiac muscle contraction} ({\tt GO:0060048}) processes occur in
a rhythmic pattern. In {\em Cardiac dysrhythmia}, however, {\em
  Cardiac muscle contraction} processes occur arrhythmically, and we
consider this to be an abnormality of the regulation of {\em Cardiac
  muscle contraction}. While these abnormalities are often highly
informative in clinical diagnostics and biological investigations, we
usually lack the necessary information that is required to
automatically determine meaningful types of abnormal distribution
patterns.

A second kind of regulatory abnormalities is related to the {\em
  onset} of a process. With respect to a reference process, a
particular kind of process may be {\em delayed} ({\tt PATO:0000502})
or {\em premature} ({\tt PATO:0000694}). For example, {\em Delayed
  apoptosis} refers to an abnormality of the {\em Regulation of
  apoptosis} in which apoptosis is induced later than normal.  We use
the PATO quality {\em onset} ({\tt PATO:...}) and its children {\em
  delayed} and {\em premature} to define these types of regulatory
abnormality. Similarly, we use PATO's {\em offset ({\tt PATO:...})
  quality and its children to characterize regulatory abnormalities in
  which a process ends prematurely or too late.

Finally, a third kind of regulatory abnormality refers to abnormal
rates with respect to a participant of the process that is being
regulated. For example, a cytoplasmic flow rate can be increased or
decreased not within a single {\em cytoplasmic streaming} process but
rather the total flow rate, as a summation over all cytoplasmic
streaming processes that occur within an organism (or a particular
anatomical location), is increased or decreased.
